Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-05-31 paper/manuscript.tex
Updates from Overleaf

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/Fig2_dummy.pdf}
    \caption{Thermodynamic cycles}
    \label{fig:fig1}
\end{figure}
\todo[inline, color=green!20]{This secion needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/Fig2_dummy.pdf}
    \caption{a) single topology b) dual topology Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:topology}
\end{figure} 
The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in which functional groups are present; e.g. consider a transformation of benzene (A) to benzyl alcohol (B)~\ref{fig:topology}.
In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group. Bond parameters as well as partial charges are adjusted accordingly. 
Thus in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:topology} (a). 
In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{Shirts2012}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{Mobley:2014:J.Comput.AidedMol.Des.}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:topology} (b) depicts how such a dual topology works.

@@

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/Fig2_dummy.pdf}
    \caption{a) restrictive MCSS match b) ring breaking}
    \label{fig:mcss}
\end{figure}

@@

Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn. These additional observables include, but are not
limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions. 
Each of these metrics have demonstrated promising results for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimate. 
Results obtained from calculations with convergence issues should be checked for errors or run for longer before any confidence should be placed in conclusions drawn from their analysis.
In relative calculations that share similar binding modes, for example, and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often not necessary due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates [ref]. 
The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.

@@

\item Computing and reporting uncertainties on free energies
correlated bootstrap v. timeseries analysis
\item Quantifying standard error in dG estimate
\item Varience in dG estimate using multiple methods (TI, DEXP, IEXP, BAR, MBAR, GDEL, etc.)
\item Agreement in dG estimate when repeating calculations with different parameters (random seeds, initial configurations, forcefield, etc.)
\item Calculating differences in free energy change as a function of time
	\item Starting from beginning or end of simulation
	\item Significance of differences in midpoint estimate (High middle error: high uncertainty)
\item Ensemble method to combine uncertainties into interpretable weighted metric
	\item Simple normalized metric to determine confidence in calculation
	\item Easily interpreted by chemist/biologist when prioritizing new chemistries
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\item Other considerations for many transformations
\todo[inline]{JM: add references, maybe a diagram ?}
Cycle closure error
Additional analyses may be performed when free energy changes for multiple transformations are available and the desired free energy difference between two molecules A and B may be obtained through more than one pathway connecting. This usually applies to relative free energy calculations. 
For instance the free energy change for the transformation of molecule A into molecule B should be the opposite of the free energy change for the transformation of molecule B into molecule A. Significant deviations from this may indicate insufficient configurational sampling along the lambda schedule of one or both transformations. The approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed network should be zero. In practice this is not observed and deviations become increasingly large as the size of the cycle increases owing to propagation of statistical errors. Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly. This may be done by extending simulations duration, or by averaging free energy changes over multiple repeats. The later approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates. 
A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change. The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors (REF). An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node (REF).

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{free_energy_trajectories.pdf}
    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e. the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3. Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data. The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3. Longer correlation times in CB8-G3 cause the calculation to converge more slowly. The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
}
    \label{fig:freeenergytrajectories}
\end{figure}

>>>>

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/Fig1_dummy.pdf}
    \caption{Thermodynamic cycles}
    \label{fig:fig1}
\end{figure}
\todo[inline, color=green!20]{This secion needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig4/topologies.pdf}
    \caption{a) single topology b) dual topology Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:topology}
\end{figure} 
The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in which functional groups are present; e.g. consider a transformation of benzene (A) to benzyl alcohol (B)~\ref{fig:topology}.
In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group. Bond parameters as well as partial charges are adjusted accordingly. 
Thus in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:topology} (a). 
In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{Shirts2012}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{Mobley:2014:J.Comput.AidedMol.Des.}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:topology} (b) depicts how such a dual topology works.

@@

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig5/MCS.pdf}
    \caption{a) restrictive MCSS match b) ring breaking}
    \label{fig:mcss}
\end{figure}

@@

Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn. These additional observables include, but are not
limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions (figure \ref{fig:convergence_forward_reverse}).

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{paper/figures/figX/Forward_Reverse_Fig.pdf}
    \caption{Free energy calculated in k$_{B}$T for two different RBFE perturbations. Each plot shows the change in free energy calculated using different fractions of simulation time (5 ns total). The top panel shows the three-step protocol for a perturbation involving 5 perturbed atoms, while the bottom panel shows the same protocol for a perturbation involving 10 perturbed atoms. The difference in energy between the forward (blue) and reverse (red) free energy calculations at the midpoint of the simulation time gives an indication of the overall convergence of the simulation, with differences over 1 k$_{B}$T indicating poor convergence.}
    \label{fig:convergence_forward_reverse}
\end{figure}

Each of these metrics have demonstrated promising results for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimate. 
Results obtained from calculations with convergence issues should be checked for errors or run for longer before any confidence should be placed in conclusions drawn from their analysis.
In relative calculations that share similar binding modes, for example, and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often not necessary due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates [ref]. 
The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.

@@

\subsubsection{Uncertainty estimation}
Analysis of independent samples of energy data from a series of equilibrium simulations allow for the estimation of the free energy and error associated with its estimate. 
Quantification of different error metrics depend on the form of analysis and method being used, such as those outlined above.
The computation of free energies using TI (section \ref{sec:step3}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of lambda states, which is required to minimize the variance in the free energy estimate. 
The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling is necessary to converge the best approximation of the integral.
Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently. 
Therefore, $var$($\Delta$G) = $\sum_{i=1}^{K}w_{k}^2 var(\frac{dU}{d\lambda})_{k}$.
For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\lambda}}$, a relatively small number of lambda states is necessary for sufficient accuracy and low variance in the free energy estimate. 
However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more lambda intermediate states become necessary to reduce error. 
It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
Compared with TI, the MBAR method (section \ref{sec:step3}) discussed above provides uncertainty estimation directly from solving complex systems of linear equations to compute the variances between all states. 
The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate. 
Additionally, agreement in the free energy across different estimators can be compared and variance computed across estimates used to assess convergence.
For example, variance in free energy estimates computed with BAR, MBAR, and TI correlated with larger errors in a set of MCL1 inhibitors.
Those perturbations where free energies across different estimators shared high agreement with each other (variance less than 0.1 kcal/mol) had a r$^2$ of 0.623 and mean unsigned error (MUE) of 0.787 kcal/mol (N=17) compared with a r$^2$ of 0.295 and MUE of 1.49 kcal/mol (N=23) (unpublished results).
Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine. 
The assessment of variability in free energy calculations due to repeating simulations has been previously reported \cite{aldeghi2019accurate}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence or problems with system set up. 
System set up errors include poor ligand or protein preparation, atom mapping, or forcefield parameters as described above.
For relative binding free energy calculations, additional sensitivity analysis can be performed by changing the initial configurations of non-core regions of the perturbation topology and determining if the proposed change results in a large differences in the computed relative free energy.
The proposed changes in configuration are increasingly relevant if no experimental evidence is available to reduce uncertainty in where the changing atoms should be positioned.
Rerunning calculations with different protein, water, or small molecule force fields (assuming they are sufficiently accurate) can also be used as an additional line of evidence that the free energy you are calculating does not change significantly based on the force field parameters utilized.

\begin{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\end{itemize}
\item Other considerations for many transformations
\todo[inline]{JM: add references, maybe a diagram ?}
Cycle closure error
Additional analyses may be performed when free energy changes for multiple transformations are available and the desired free energy difference between two molecules A and B may be obtained through more than one pathway connecting. This usually applies to relative free energy calculations. 
For instance the free energy change for the transformation of molecule A into molecule B should be the opposite of the free energy change for the transformation of molecule B into molecule A. Significant deviations from this may indicate insufficient configurational sampling along the lambda schedule of one or both transformations. The approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed network should be zero. In practice this is not observed and deviations become increasingly large as the size of the cycle increases owing to propagation of statistical errors. Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly. This may be done by extending simulations duration, or by averaging free energy changes over multiple repeats. The later approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates. 
A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change. The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors (REF). An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node (REF).

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{fig6/free_energy_trajectories.pdf}
    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e. the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3. Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data. The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3. Longer correlation times in CB8-G3 cause the calculation to converge more slowly. The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
}
    \label{fig:freeenergytrajectories}
\end{figure}
